The Forecasting Trap -- notes

### 2022-02-04 notebook/alternate-models.Rmd
WIP :construction: alternate examples

    }
    df <- tibble(t = 1:Tmax, x1, x2, x3, x4, x5)
    df
}

@@

```{r}
clip <- function(x, lower=0, upper=1) {
    x[x <= lower] <- lower
    x[x >= upper] <- upper
    x
}

utility <- function(x) {
    delta <- 0.999
    w3 <- 0.5
    w2 <- 0.25
    w1 <- 1 - (w2 + w3)
    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    df <- model_5sp(u_prey, u_pred)
    R <- 
        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
        w2 * u_pred * df$x4 +
        w3 * df$x5
    t <- seq_along(R)
    - as.numeric(R %*% delta ^ t)
    }

-utility(c(0.5,0.5))
-utility(c(0,1))
-utility(c(1,0))
-utility(c(0,0))

@@

o <- optim(c(0.5,0.5), utility, method="SANN")
clip(o$par, 0, 1)
o$value

@@

df %>%
    pivot_longer(tidyselect::starts_with("x"),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(t, biomass, col= species)) + geom_path()
```

@@

```{r}
df <- model_5sp(0, 0., sigma = 0.01)

df %>%
    transmute(time = t, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
    pivot_longer(any_of(c("cormorant", "bass", "herring")),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(time, biomass, col= species)) + geom_path()
```

A reasonable single-species model for Bass achieves the optimal solution.
The model is reasonable in the sense that it is a simple Ricker model 
that comparably reproduces the unfished dynamics of the Bass.
Nevertheless, it is actually quite poor at forecasting fished stock sizes.

@@

model_1sp <- function(u = 0, A = 2, R = 1.6, Tmax=50, x1o = .8, sigma = 0.01) {

    omega1 <- rnorm(Tmax, 0, sigma)

    s1 <- x1 <- numeric(Tmax)
    x1[1] <- x1o

    for (t in 1:(Tmax-1)) {
        s1[t] <- x1[t] * (1 - u)
        x1[t+1] <- s1[t] * exp(R - s1[t] * A + omega1[t])
    }

    df <- tibble(time = 1:Tmax, x1)
}

df <- model_1sp(0)
df %>% ggplot(aes(t, x1)) + geom_path()

@@

```{r}
utility <- function(x) {
    delta <- 0.999
    u <- clip(x, 0, 1)
    df <- model_1sp(u)
    R <- u * df$x1 
    t <- seq_along(R)
    - as.numeric(R %*% delta ^ t)
    }

o <- optimize(utility, c(0,1))
o$minimum
```

```{r}
df <- model_1sp(o$minimum)
df %>% ggplot(aes(t, x1)) + geom_path()
```

```{r}
R1 <- R2 <- 3
A1 <- A2 <- 2
B1 <- 0.5
B2 <- -0.3
Tmax <- 50
omega1 <- rnorm(Tmax, 0, 0.001)
omega2 <- rnorm(Tmax, 0, 0.001)

u1 <- u2 <- s1 <- s2 <- x1 <- x2 <- numeric(Tmax)
x1[1] <- 1
x2[1] <- 1

# or, constant strategy
u1 <- rep(.2, Tmax)
u2 <- rep(.5, Tmax)

for (t in 1:(Tmax-1)) {
    s1[t] <- x1[t] * (1 - u1[t])
    s2[t] <- x2[t] * (1 - u2[t])
    x1[t+1] <- s1[t] * exp(R1 - A1 * s1[t] - B1 * s2[t] + omega1[t])
    x2[t+1] <- s2[t] * exp(R2 - A2 * s1[t] - B2 * s1[t] + omega2[t])
}

df <- data.frame(time = 1:Tmax, x1, x2, u1, u2)
```

>>>>

    }
    df <- tibble(time = 1:Tmax, x1, x2, x3, x4, x5)
    df
}

@@

```{r}
clip <- function(x, lower=0, upper=1) {
    x[x <= lower] <- lower
    x[x >= upper] <- upper
    x
}
```

reward <- function(df,
                   x,
                   w3 = 0.5,
                   w2 = 0.25,
                   w1 = 1 - (w2+w3),
                   delta = 0.999
                  ) {
    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    R <-
        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
        w2 * u_pred * df$x4 +
        w3 * df$x5
    t <- seq_along(R)
    as.numeric(R %*% delta ^ t)
}

utility <- function(x, f = model_5sp) {

    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    ## ought to average over many replicates!
    df <- f(u_prey, u_pred)

    # negative since optimizer minimizes
    - reward(df, x)
}

```{r}
-utility(c(0.5,0.5))
-utility(c(0,1))
-utility(c(1,0))
-utility(c(0,0))

@@

#o <- optim(c(0.5,0.5), utility, method="SANN")
#clip(o$par, 0, 1)
#o$value

@@

df %>%
    pivot_longer(tidyselect::starts_with("x"),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(time, biomass, col= species)) + geom_path()
```

@@

```{r}
## Three-species 'view' of the underlying dynamics, unfished

df <- model_5sp(0, 0., sigma = 0.01, x1o =.1, x2o = .1, x3o = .1, x4o = .1, x5o = .1)
df %>%
    transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
    pivot_longer(any_of(c("cormorant", "bass", "herring")),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(time, biomass, col= species)) + geom_path()
```

Now consider an alternative model which ignores some of this biological complexity:
Our alternate model has several oversimplifications:
(1) it treat all three herring species as a single species,
(2) it fails to capture the coupled predator-prey dynamics between bass and herring, 
(3) it oversimplifies the cormorant dynamics, assuming the cormorant population is determined to be a fixed fraction of the herring,
and ignoring the impact of cormorant's predation on the herring itself.
(4) Lastly, our model will overestimate the mortality introduced on herring by a given fishing effort.
These elements are all obviously wrong, but not so arbitrary as to be inconceivable as a candidate model.
Researchers frequently consider models which make oversimplifications all the time, and rely on model choice processes to weed them out.

The parameterization chosen for the oversimplified 3 species model can reasonably reproduce the unfished dynamics.
Under any harvesting regime or other influence that perturbs the system significantly far from the unfished equilibrium co-existence state of the model quickly reveals the poor forecasting ability of this model.
Nevertheless, solving for the optimal policy (under the same constraints of constant harvest fractions as we consider for the true model) results in a harvest policy which provides nearly optimal performance.

```{r}
model_3sp <- function(u_prey = 0,
                      u_predator = 0,
                      A1 = 2,
                      R1 = 1.6,
                      A2 = 2.75,
                      R2 = 1.6,
                      Tmax = 50,
                      x1o = .8, # Herring
                      x2o = .6, # Bass
                      x3o = 0.15, # Cormorant
                      sigma = 0.01) {

    omega1 <- rnorm(Tmax, 0, sigma)
    omega2 <- rnorm(Tmax, 0, sigma)
    omega3 <- rnorm(Tmax, 0, sigma)

    s1 <- x1 <- numeric(Tmax)
    s2 <- x2 <- numeric(Tmax)
    x3 <- numeric(Tmax)
    x1[1] <- x1o
    x2[1] <- x2o
    x3[1] <- x3o

    for (t in 1:(Tmax-1)) {
        s1[t] <- x1[t] * (1 - 10*u_prey)
        s2[t] <- x2[t] * (1 - u_predator)

        x1[t+1] <- s1[t] * exp(R1 - s1[t] * A1 + omega1[t])
        x2[t+1] <- s2[t] * exp(R2 - s2[t] * A2 + omega2[t])
        x3[t+1] <- .25 * x1[t]
    }

    df <- tibble(time = 1:Tmax, x1, x4 = x2, x5 = x3, x2 = 0, x3 = 0)
}
```

```{r}
# unfished equilibrium
df <- model_3sp(0, 0)
df %>%
    transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
    pivot_longer(any_of(c("cormorant", "bass", "herring")),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(time, biomass, col= species)) + geom_path()
```

```{r}
# Optimal policy under identical utility
u_3sp <- function(x) utility(x, f = model_3sp)

m3sp <- optim(c(0.5,0.1), u_3sp)
m3sp$value
clip(m3sp$par, 0, 1)

Alternate model, better forecast, worse decision:

```{r}
model_B <- function(u_prey = 0.5,
                      u_pred = 0.5,
                      sigma = 0.05,
                      x1o = 0.796, # unfushed equib
                      x4o = 0.600,
                      x5o = 0.280,
                      Tmax = 50,
                      R1 = 1.0213,
                      A11 = 1,
                      A14 = 0.7252,
                      A15 = 0.2445,
                      A41 = 0.7252,
                      A51 = 0.2445) {
    x1 <- x4 <- x5 <- numeric(Tmax)
    s1 <-  s4 <- numeric(Tmax)
    xi1 <- rnorm(Tmax, 0, sigma)
    xi4 <- rnorm(Tmax, 0, sigma)
    xi5 <- rnorm(Tmax, 0, sigma)

    x1[1] <- x1o
    x4[1] <- x4o
    x5[1] <- x5o

    for (t in 1:(Tmax-1)) {
        s1[t] <- x1[t] * (1 - u_prey)
        s4[t] <- x4[t] * (1 - u_pred)
        x1[t+1] = s1[t]*exp(R1 - A11*s1[t] - A14*s4[t] - A15 * x5[t] + xi1[t])
        x4[t+1] = s4[t]*exp(A41*s1[t] - s4[t] + xi4[t])
        x5[t+1] = x5[t]*exp(A51*s1[t] - x5[t] +  xi5[t])

    }
    df <- tibble(time = 1:Tmax, x1, x4, x5, x2 =0, x3 = 0)
    df
}
```

```{r}
# Optimal policy under identical utility
u_B <- function(x) utility(x, f = model_B)
u_B(c(0, .63))
#optim(c(0.5,0.1), u_B, method = "L-BFGS-B", lower=0, upper=1)

B <- optim(c(0.5,0.1), u_B)
B$value
clip(B$par, 0, 1)

@@

utility( c(clip(o$par[[1]]), clip(o$par[[2]])) )

utility( c(clip(m3sp$par[[1]]), clip(m3sp$par[[2]])) )

utility( c(clip(B$par[[1]]), clip(B$par[[2]])) )

@@

Run the simulation using management policy of each

## Generate replicate predictions from each model at each step.
## compute CRPS of an observation from the "true" model vs those distributions

function(est, f) {
    u_prey <- clip(est$par[[1]])
    u_pred <- clip(est$par[[2]])
    map_dfr(1:100, function(i) {
    df <- f(u_prey, u_pred, x1o = .1, x2o = .1, x3o = .1, x4o = .1, x5o = .1)
    df %>%
        transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
        pivot_longer(any_of(c("cormorant", "bass", "herring")),
                        names_to = "species",
                        values_to = "biomass")
    }, .id = "rep")

}

```{r}
# management of true model under optimal pars for model_3sp
df <- model_5sp(clip(m3sp$par[[1]],0,1),
                clip(m3sp$par[[2]],0,1),
                 x1o = .1, x4o = .1, x5o = .1)
df %>%
    transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3) %>%
    pivot_longer(any_of(c("cormorant", "bass", "herring")),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(time, biomass, col= species)) + geom_path()
```
